---
title: "project_2"
output: html_document
date: "2026-07-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# reticulate is used to read the .pkl files directly (they were written from
# Python/pickle). py_load_object() reads a pickle file and converts it to a
# native R object (dict -> named list, numpy array -> R array).
library(reticulate)

# Helper: plot one discrete-state frame the way matplotlib's imshow does.
# `mat` is a 2D matrix of integer codes (e.g. 0/1 or 0/1/2/3).
# `cols` is a vector of colors, one per code, indexed from the minimum code.


plot_frame <- function(mat, cols, main = "") {
  n <- length(cols)
  # image() plots row 1 at the bottom and transposes rows/cols relative to
  # how the matrix looks when printed, so flip/transpose to match imshow.
  mat_img <- t(mat[nrow(mat):1, ])
  image(
    mat_img,
    col = cols,
    breaks = seq(-0.5, n - 0.5, by = 1),
    zlim = c(0, n - 1),
    axes = FALSE, main = main, asp = 1
  )
}

extract_frame <- function(A_list, i) {
  if (is.list(A_list) && !is.array(A_list)) {
    mat <- A_list[[i]]
  } else {
    mat <- A_list[i, , ]
  }
  as.matrix(mat)
}


number_of_cells
```


# Project 2: determining effect of drug on cell-to-cell virus transmission

#### dataset 2.1: virus spreading with no drug added

infection_no_drug_experiment_0.pkl,
infection_no_drug_experiment_1.pkl,
infection_no_drug_experiment_2.pkl

Cell centroids are again saved in `A_list`, which has shape (t, x, y). For
each time point t, there is a spatial field of cells in the xy-plane, where
each value encodes the cell's state: 0 = no cell, 1 = susceptible cell,
2 = infected cell, 3 = recovered cell. For example, `A_list[t, x, y] == 2`
means the cell at position (x, y) is infected at frame t.

See metadata for spatial length scale and interval between time points.

The biologists collected 3 movies (experiments 0, 1, and 2).

```{r load-plot-infection-example}
# example of loading and plotting data:

output <- py_load_object("infection_no_drug_experiment_2.pkl")

A_list   <- output$A_list
metadata <- output$metadata

cat("Size of data (t, 2D position): ", paste(dim(A_list), collapse = " x "), "\n")
cat("\nData Details:\n")
for (key in names(metadata)) {
  cat(key, "-", as.character(metadata[[key]]), "\n")
}


unlist(lapply((lapply(A_list, function(x) apply(x, 1, function(y) sum(y > 0) ))), sum))

# discrete colormap: one color per state (0=empty, 1=S, 2=I, 3=R)
cmap <- c(
  "white",                    # 0 empty
  "blue",                     # 1 susceptible
  rgb(0, 200 / 256, 0),       # 2 infected
  "black"                     # 3 recovered
)


times_to_plot <- c(0, 40, 400)   # 0-indexed, as in the Python version

par(mfrow = c(1, length(times_to_plot)), mar = c(1, 1, 2, 1))
for (t in times_to_plot) {
 plot_frame(extract_frame(A_list, t + 1), cols = cmap, main = paste("t =", t))
}
legend("topright", inset = c(-0.3, 0),
       legend = c("Susceptible", "Infected", "Recovered"),
       fill = cmap[2:4], xpd = TRUE, bty = "n")
```

#### dataset 2.2: virus spreading with drug added

infection_with_drug_experiment_0.pkl
infection_with_drug_experiment_1.pkl
infection_with_drug_experiment_2.pkl

Cell centroids are again saved in `A_list`, which has shape (t, x, y). For
each time point t, there is a spatial field of cells in the xy-plane, where
each value encodes the cell's state: 0 = no cell, 1 = susceptible cell,
2 = infected cell, 3 = recovered cell. For example, `A_list[t, x, y] == 2`
means the cell at position (x, y) is infected at frame t.

See metadata for spatial length scale and interval between time points.

The biologists collected 3 movies (experiments 0, 1, and 2).


```{r load-plot-infection-example}
# example of loading and plotting data:

output <- py_load_object("infection_with_drug_experiment_2.pkl")

A_list   <- output$A_list
metadata <- output$metadata

cat("Size of data (t, 2D position): ", paste(dim(A_list), collapse = " x "), "\n")
cat("\nData Details:\n")
for (key in names(metadata)) {
  cat(key, "-", as.character(metadata[[key]]), "\n")
}


unlist(lapply((lapply(A_list, function(x) apply(x, 1, function(y) sum(y > 0) ))), sum))

# discrete colormap: one color per state (0=empty, 1=S, 2=I, 3=R)
cmap <- c(
  "white",                    # 0 empty
  "blue",                     # 1 susceptible
  rgb(0, 200 / 256, 0),       # 2 infected
  "black"                     # 3 recovered
)


times_to_plot <- c(0, 1, 2)   # 0-indexed, as in the Python version

par(mfrow = c(1, length(times_to_plot)), mar = c(1, 1, 2, 1))
for (t in times_to_plot) {
 plot_frame(extract_frame(A_list, t + 1), cols = cmap, main = paste("t =", t))
}
legend("topright", inset = c(-0.3, 0),
       legend = c("Susceptible", "Infected", "Recovered"),
       fill = cmap[2:4], xpd = TRUE, bty = "n")
```

```{r}
library(clue)   # solve_LSAP = Hungarian algorithm

link_frames <- function(frame_t, frame_t1, max_dist) {
  n1 <- nrow(frame_t); n2 <- nrow(frame_t1)
  D  <- as.matrix(dist(rbind(frame_t[, c("x","y")], frame_t1[, c("x","y")])))
  D  <- D[1:n1, (n1 + 1):(n1 + n2), drop = FALSE]

  cost <- D
  cost[cost > max_dist] <- 1e6   # disallow biologically implausible jumps

  assignment <- solve_LSAP(cost)     # optimal one-to-one match, min total distance
  idx_t1     <- as.integer(assignment)
  dist_val   <- D[cbind(seq_len(n1), idx_t1)]
  idx_t1[dist_val > max_dist] <- NA   # treat as lost/new cell instead of forcing a bad match

  data.frame(idx_t = seq_len(n1), idx_t1 = idx_t1, dist = dist_val)
}
```

