---
title: "project_2"
output: html_document
date: "2026-07-06"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# reticulate is used to read the .pkl files directly (they were written from
# Python/pickle). py_load_object() reads a pickle file and converts it to a
# native R object (dict -> named list, numpy array -> R array).
library(reticulate)

# Helper: plot one discrete-state frame the way matplotlib's imshow does.
# `mat` is a 2D matrix of integer codes (e.g. 0/1 or 0/1/2/3).
# `cols` is a vector of colors, one per code, indexed from the minimum code.


plot_frame <- function(mat, cols, main = "") {
  n <- length(cols)
  # image() plots row 1 at the bottom and transposes rows/cols relative to
  # how the matrix looks when printed, so flip/transpose to match imshow.
  mat_img <- t(mat[nrow(mat):1, ])
  image(
    mat_img,
    col = cols,
    breaks = seq(-0.5, n - 0.5, by = 1),
    zlim = c(0, n - 1),
    axes = FALSE, main = main, asp = 1
  )
}

extract_frame <- function(A_list, i) {
  if (is.list(A_list) && !is.array(A_list)) {
    mat <- A_list[[i]]
  } else {
    mat <- A_list[i, , ]
  }
  as.matrix(mat)
}


number_of_cells
```


# Project 2: determining effect of drug on cell-to-cell virus transmission

#### dataset 2.1: virus spreading with no drug added

infection_no_drug_experiment_0.pkl,
infection_no_drug_experiment_1.pkl,
infection_no_drug_experiment_2.pkl

Cell centroids are again saved in `A_list`, which has shape (t, x, y). For
each time point t, there is a spatial field of cells in the xy-plane, where
each value encodes the cell's state: 0 = no cell, 1 = susceptible cell,
2 = infected cell, 3 = recovered cell. For example, `A_list[t, x, y] == 2`
means the cell at position (x, y) is infected at frame t.

See metadata for spatial length scale and interval between time points.

The biologists collected 3 movies (experiments 0, 1, and 2).

```{r load-plot-infection-example}
# example of loading and plotting data:

output <- py_load_object("infection_no_drug_experiment_2.pkl")

A_list   <- output$A_list
metadata <- output$metadata

cat("Size of data (t, 2D position): ", paste(dim(A_list), collapse = " x "), "\n")
cat("\nData Details:\n")
for (key in names(metadata)) {
  cat(key, "-", as.character(metadata[[key]]), "\n")
}


unlist(lapply((lapply(A_list, function(x) apply(x, 1, function(y) sum(y > 0) ))), sum))

# discrete colormap: one color per state (0=empty, 1=S, 2=I, 3=R)
cmap <- c(
  "white",                    # 0 empty
  "blue",                     # 1 susceptible
  rgb(0, 200 / 256, 0),       # 2 infected
  "black"                     # 3 recovered
)


times_to_plot <- c(0, 40, 400)   # 0-indexed, as in the Python version

par(mfrow = c(1, length(times_to_plot)), mar = c(1, 1, 2, 1))
for (t in times_to_plot) {
 plot_frame(extract_frame(A_list, t + 1), cols = cmap, main = paste("t =", t))
}
legend("topright", inset = c(-0.3, 0),
       legend = c("Susceptible", "Infected", "Recovered"),
       fill = cmap[2:4], xpd = TRUE, bty = "n")
```

#### dataset 2.2: virus spreading with drug added

infection_with_drug_experiment_0.pkl
infection_with_drug_experiment_1.pkl
infection_with_drug_experiment_2.pkl

Cell centroids are again saved in `A_list`, which has shape (t, x, y). For
each time point t, there is a spatial field of cells in the xy-plane, where
each value encodes the cell's state: 0 = no cell, 1 = susceptible cell,
2 = infected cell, 3 = recovered cell. For example, `A_list[t, x, y] == 2`
means the cell at position (x, y) is infected at frame t.

See metadata for spatial length scale and interval between time points.

The biologists collected 3 movies (experiments 0, 1, and 2).


```{r load-plot-infection-example}
# example of loading and plotting data:

output <- py_load_object("/Applications/mism-hackathon/data/raw/infection_with_drug_experiment_0.pkl")

A_list   <- output$A_list
metadata <- output$metadata

cat("Size of data (t, 2D position): ", paste(dim(A_list), collapse = " x "), "\n")
cat("\nData Details:\n")
for (key in names(metadata)) {
  cat(key, "-", as.character(metadata[[key]]), "\n")
}


unlist(lapply((lapply(A_list, function(x) apply(x, 1, function(y) sum(y > 0) ))), sum))

# discrete colormap: one color per state (0=empty, 1=S, 2=I, 3=R)
cmap <- c(
  "white",                    # 0 empty
  "blue",                     # 1 susceptible
  rgb(0, 200 / 256, 0),       # 2 infected
  "black"                     # 3 recovered
)


times_to_plot <- c(0, 1, 2)   # 0-indexed, as in the Python version

par(mfrow = c(1, length(times_to_plot)), mar = c(1, 1, 2, 1))
for (t in times_to_plot) {
 plot_frame(extract_frame(A_list, t + 1), cols = cmap, main = paste("t =", t))
}
legend("topright", inset = c(-0.3, 0),
       legend = c("Susceptible", "Infected", "Recovered"),
       fill = cmap[2:4], xpd = TRUE, bty = "n")
```



```{r}
tracks_nodrug = read.csv("/Applications/mism-hackathon/project/cell-infection/results/trajectory_no_drug_0.csv")
tracks
```


## Having the dataset for all 6 conditions


```{r}
tracks <- list()

results_dir <- file.path(script_dir, "results")
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

for (cond in conditions) {
  for (e in experiments) {

    ## -----------------------
    ## Load data
    ## -----------------------
    f   <- file.path(DATA,
                     sprintf("infection_%s_experiment_%d.pkl", cond, e))

    out <- py_load_object(f)
    A   <- out$A_list

    L <- as.integer(out$metadata[["xy dimensions"]])
    if (length(L) == 0 || is.na(L))
      L <- nrow(get_frame(A, 1))

    dt <- as.numeric(out$metadata[["time interval (mins)"]])

    ## -----------------------
    ## Track cells
    ## -----------------------
    tr <- build_tracks(
      A,
      L,
      periodic = TRUE,
      solver = fast_solver,
      state = NULL
    )

    d <- diagnostics(tr, L, periodic = TRUE)

    step <- sqrt(tr$disp[, , 1]^2 + tr$disp[, , 2]^2)

    frac_still <- mean(step == 0, na.rm = TRUE)
    speed <- d$mean_step / dt

    key <- paste(cond, e, sep = "_")
    tracks[[key]] <- tr

    cat(sprintf("%-10s %4d %6d %11.3f %11.3f %8.3f %9.3f %7.0f%%\n",
                cond, e,
                d$n_cells,
                d$mean_step,
                d$nn_spacing,
                d$rho,
                speed,
                100 * frac_still))

    ## -----------------------
    ## Save trajectory plot
    ## -----------------------
    plot_tracks(
      tr,
      L = L,
      main = sprintf("Infection assay: %s (exp %d)", cond, e),
      file = file.path(
        results_dir,
        sprintf("tracks_infection_%s_%d.png", cond, e)
      )
    )

    ## -----------------------
    ## Save trajectory dataframe
    ## -----------------------
    traj <- tracks_to_df(
      tr,
      dt = dt,
      A = A
    )

    write.csv(
      traj,
      file.path(
        results_dir,
        sprintf("trajectory_%s_%d.csv", cond, e)
      ),
      row.names = FALSE
    )

    ## -----------------------
    ## Save one frame-to-frame link
    ## -----------------------
    plot_link(
      tr,
      t = 1,
      file = file.path(
        results_dir,
        sprintf("link_%s_%d_frame1.png", cond, e)
      )
    )

    ## -----------------------
    ## Console summary
    ## -----------------------
    cat(sprintf(
      "\nSaved outputs for %s experiment %d\n",
      cond, e
    ))

    cat(sprintf(
      "Trajectory: %d rows (%d cells × %d frames)\n",
      nrow(traj),
      length(unique(traj$cell)),
      length(unique(traj$frame))
    ))

    print(table(traj$state_label))
    cat("\n--------------------------------------------\n")
  }
}

cat("\nFinished processing all six infection datasets.\n")
```


```{r}
## =========================================================================
## Spatial SIR hazard model for cell-tracking data (before/after drug)
## =========================================================================
## Expected input: two data.frames, each with columns
##   cell        : cell ID (1..161)
##   frame       : 1..401
##   t_min       : seq(0, 200, 0.5)
##   x, y        : coordinates
##   state       : 1 = S, 2 = I, 3 = R
##   state_label : text label (not used in fitting)
##
## Model:
##   lambda_i(t) = beta1 * exp(betaD * drug) * sum_{j in I(t)} exp(-d_ij(t)/phi)
##   P(S_i -> I_i in [t, t+delta)) = 1 - exp(-lambda_i(t) * delta)
##   P(I_i -> R_i in [t, t+delta)) = 1 - exp(-gamma * exp(gammaD * drug) * delta)
## =========================================================================

library(dplyr)
library(bbmle)

## -------------------------------------------------------------------------
## 0. Load and stack the two datasets
## -------------------------------------------------------------------------
# df_before <- read.csv("before_drug.csv")
# df_after  <- read.csv("after_drug.csv")

df_before$drug <- 0
df_after$drug  <- 1

tracks <- bind_rows(df_before, df_after) %>%
  arrange(drug, cell, frame)

delta <- 1   # 1 frame = 0.5 min; hazard/rate parameters will be in units of "per frame"

## -------------------------------------------------------------------------
## 1. Precompute, for every (drug condition, frame), the distance matrix
##    between susceptible cells and infected cells, plus the outcome
##    (did each susceptible cell become infected by the next frame?)
##    Distances don't depend on model parameters, so this is done ONCE.
## -------------------------------------------------------------------------
build_transmission_blocks <- function(tracks, delta = 1) {
  blocks <- list()
  k <- 1

  for (d in unique(tracks$drug)) {
    sub <- tracks %>% filter(drug == d)
    frames <- sort(unique(sub$frame))

    for (t in frames) {
      if (!(t + delta) %in% frames) next

      now <- sub %>% filter(frame == t)
      nxt <- sub %>% filter(frame == t + delta) %>% select(cell, state_next = state)

      susc <- now %>% filter(state == 1) %>% left_join(nxt, by = "cell")
      inf  <- now %>% filter(state == 2)

      if (nrow(susc) == 0 || nrow(inf) == 0) next

      D <- sqrt(outer(susc$x, inf$x, "-")^2 + outer(susc$y, inf$y, "-")^2)
      # D: rows = susceptible cells, cols = currently infected cells

      blocks[[k]] <- list(
        D = D,
        became_infected = as.numeric(susc$state_next == 2),
        drug = d
      )
      k <- k + 1
    }
  }
  blocks
}

## -------------------------------------------------------------------------
## 2. Recovery data (I -> R), no distance needed
## -------------------------------------------------------------------------
build_recovery_data <- function(tracks, delta = 1) {
  out <- list()
  k <- 1

  for (d in unique(tracks$drug)) {
    sub <- tracks %>% filter(drug == d)
    frames <- sort(unique(sub$frame))

    for (t in frames) {
      if (!(t + delta) %in% frames) next

      now <- sub %>% filter(frame == t, state == 2)
      nxt <- sub %>% filter(frame == t + delta) %>% select(cell, state_next = state)
      if (nrow(now) == 0) next

      now <- now %>% left_join(nxt, by = "cell")
      out[[k]] <- data.frame(
        became_removed = as.numeric(now$state_next == 3),
        drug = d
      )
      k <- k + 1
    }
  }
  bind_rows(out)
}

trans_blocks <- build_transmission_blocks(tracks, delta)
recov_df     <- build_recovery_data(tracks, delta)

cat("Number of (frame, condition) transmission blocks:", length(trans_blocks), "\n")
cat("Number of recovery observations:", nrow(recov_df), "\n")

## -------------------------------------------------------------------------
## 3. Negative log-likelihood
## -------------------------------------------------------------------------
nll <- function(log_beta1, log_phi, betaD, log_gamma, gammaD) {

  beta1 <- exp(log_beta1)
  phi   <- exp(log_phi)
  gamma <- exp(log_gamma)

  ## --- transmission log-likelihood ---
  ll_trans <- 0
  for (blk in trans_blocks) {
    kappa_sum <- rowSums(exp(-blk$D / phi))
    lambda    <- beta1 * exp(betaD * blk$drug) * kappa_sum
    p_inf     <- 1 - exp(-lambda * delta)
    p_inf     <- pmin(pmax(p_inf, 1e-10), 1 - 1e-10)
    ll_trans  <- ll_trans + sum(dbinom(blk$became_infected, 1, p_inf, log = TRUE))
  }

  ## --- recovery log-likelihood ---
  p_rec <- 1 - exp(-gamma * exp(gammaD * recov_df$drug) * delta)
  p_rec <- pmin(pmax(p_rec, 1e-10), 1 - 1e-10)
  ll_rec <- sum(dbinom(recov_df$became_removed, 1, p_rec, log = TRUE))

  -(ll_trans + ll_rec)
}

## -------------------------------------------------------------------------
## 4. Fit
## -------------------------------------------------------------------------
## Reasonable starting values: eyeball beta1 from crude early growth rate,
## phi from grid spacing (e.g. ~ a few cell-widths), gamma from mean
## infectious-period length. Adjust these if optimizer struggles.
start_vals <- list(
  log_beta1 = log(0.01),
  log_phi   = log(3),
  betaD     = 0,
  log_gamma = log(0.02),
  gammaD    = 0
)

fit <- mle2(nll, start = start_vals, method = "BFGS",
            control = list(maxit = 500))

summary(fit)
ci <- confint(fit)
print(ci)

## -------------------------------------------------------------------------
## 5. Interpret results
## -------------------------------------------------------------------------
est <- coef(fit)

cat("\n--- Rate of infection ---\n")
cat("beta1 (baseline transmission rate, per frame):", exp(est["log_beta1"]), "\n")

cat("\n--- Spatial range of infection ---\n")
cat("phi (kernel decay distance, grid units):", exp(est["log_phi"]), "\n")

cat("\n--- Drug effect on transmission ---\n")
cat("betaD:", est["betaD"],
    " (exp(betaD) =", exp(est["betaD"]), "x change in transmission rate)\n")
cat("95% CI for betaD:", ci["betaD", ], "\n")
cat("--> CI excludes 0:", !(ci["betaD", 1] < 0 & ci["betaD", 2] > 0), "\n")

cat("\n--- Drug effect on recovery ---\n")
cat("gammaD:", est["gammaD"],
    " (exp(gammaD) =", exp(est["gammaD"]), "x change in recovery rate)\n")
cat("95% CI for gammaD:", ci["gammaD", ], "\n")

## -------------------------------------------------------------------------
## 6. Visual check: fitted kernel shape
## -------------------------------------------------------------------------
library(ggplot2)

d_seq <- seq(0, 20, length.out = 200)
kernel_df <- data.frame(
  d = d_seq,
  kappa = exp(-d_seq / exp(est["log_phi"]))
)

ggplot(kernel_df, aes(d, kappa)) +
  geom_line() +
  labs(x = "Distance between cells (grid units)",
       y = "Relative transmission weight",
       title = "Fitted spatial transmission kernel") +
  theme_minimal()

## -------------------------------------------------------------------------
## 7. Posterior predictive check: simulate and compare epidemic curves
## -------------------------------------------------------------------------
simulate_epidemic <- function(tracks_template, params, delta = 1) {
  # tracks_template: one condition's (cell, frame, x, y) skeleton, with
  # initial states at frame 1, used to re-simulate forward under the fit
  frames <- sort(unique(tracks_template$frame))
  state  <- tracks_template %>% filter(frame == frames[1]) %>%
    select(cell, x, y, state)

  sim_states <- list(state)

  for (t in frames[-length(frames)]) {
    coords_now <- tracks_template %>% filter(frame == t) %>% select(cell, x, y)
    state <- state %>% select(-x, -y) %>% left_join(coords_now, by = "cell")

    susc <- state %>% filter(state == 1)
    inf  <- state %>% filter(state == 2)

    if (nrow(susc) > 0 && nrow(inf) > 0) {
      D <- sqrt(outer(susc$x, inf$x, "-")^2 + outer(susc$y, inf$y, "-")^2)
      kappa_sum <- rowSums(exp(-D / params["phi"]))
      lambda <- params["beta1"] * exp(params["betaD"] * unique(tracks_template$drug)) * kappa_sum
      p_inf  <- 1 - exp(-lambda * delta)
      newly_infected <- susc$cell[runif(nrow(susc)) < p_inf]
      state$state[state$cell %in% newly_infected] <- 2
    }

    infd <- state %>% filter(state == 2)
    if (nrow(infd) > 0) {
      p_rec <- 1 - exp(-params["gamma"] * exp(params["gammaD"] * unique(tracks_template$drug)) * delta)
      newly_removed <- infd$cell[runif(nrow(infd)) < p_rec]
      state$state[state$cell %in% newly_removed] <- 3
    }

    sim_states[[length(sim_states) + 1]] <- state
    state$frame <- t + delta
  }

  bind_rows(sim_states)
}

# Example: compare observed vs simulated I(t) curve for one condition
# params <- c(beta1 = exp(est["log_beta1"]), phi = exp(est["log_phi"]),
#             betaD = est["betaD"], gamma = exp(est["log_gamma"]), gammaD = est["gammaD"])
# sim <- replicate(20, simulate_epidemic(df_before, params, delta), simplify = FALSE)
# (then plot mean +/- range of simulated I(t) against observed I(t))
```



```{r}
## =========================================================================
## Spatial SIR hazard model for cell-tracking data (before/after drug)
## =========================================================================
## Expected input: two data.frames, each with columns
##   cell        : cell ID (1..161)
##   frame       : 1..401
##   t_min       : seq(0, 200, 0.5)
##   x, y        : coordinates
##   state       : 1 = S, 2 = I, 3 = R
##   state_label : text label (not used in fitting)
##
## Model:
##   lambda_i(t) = beta1 * exp(betaD * drug) * sum_{j in I(t)} exp(-d_ij(t)/phi)
##   P(S_i -> I_i in [t, t+delta)) = 1 - exp(-lambda_i(t) * delta)
##   P(I_i -> R_i in [t, t+delta)) = 1 - exp(-gamma * exp(gammaD * drug) * delta)
## =========================================================================

library(dplyr)
library(bbmle)

## -------------------------------------------------------------------------
## 0. Load and stack the two datasets
## -------------------------------------------------------------------------
 df_before <- read.csv("/Applications/mism-hackathon/project/cell-infection/results/trajectory_no_drug_0.csv")
 df_after  <- read.csv("/Applications/mism-hackathon/project/cell-infection/results/trajectory_with_drug_0.csv")

df_before$drug <- 0
df_after$drug  <- 1

tracks <- bind_rows(df_before, df_after) %>%
  arrange(drug, cell, frame)

delta <- 1   # 1 frame = 0.5 min; hazard/rate parameters will be in units of "per frame"

## -------------------------------------------------------------------------
## 1. Precompute, for every (drug condition, frame), the distance matrix
##    between susceptible cells and infected cells, plus the outcome
##    (did each susceptible cell become infected by the next frame?)
##    Distances don't depend on model parameters, so this is done ONCE.
## -------------------------------------------------------------------------
build_transmission_blocks <- function(tracks, delta = 1) {
  blocks <- list()
  k <- 1

  for (d in unique(tracks$drug)) {
    sub <- tracks %>% filter(drug == d)
    frames <- sort(unique(sub$frame))

    for (t in frames) {
      if (!(t + delta) %in% frames) next

      now <- sub %>% filter(frame == t)
      nxt <- sub %>% filter(frame == t + delta) %>% select(cell, state_next = state)

      susc <- now %>% filter(state == 1) %>% left_join(nxt, by = "cell")
      inf  <- now %>% filter(state == 2)

      if (nrow(susc) == 0 || nrow(inf) == 0) next

      D <- sqrt(outer(susc$x, inf$x, "-")^2 + outer(susc$y, inf$y, "-")^2)
      # D: rows = susceptible cells, cols = currently infected cells

      blocks[[k]] <- list(
        D = D,
        became_infected = as.numeric(susc$state_next == 2),
        drug = d
      )
      k <- k + 1
    }
  }
  blocks
}

## -------------------------------------------------------------------------
## 2. Recovery data (I -> R), no distance needed
## -------------------------------------------------------------------------
build_recovery_data <- function(tracks, delta = 1) {
  out <- list()
  k <- 1

  for (d in unique(tracks$drug)) {
    sub <- tracks %>% filter(drug == d)
    frames <- sort(unique(sub$frame))

    for (t in frames) {
      if (!(t + delta) %in% frames) next

      now <- sub %>% filter(frame == t, state == 2)
      nxt <- sub %>% filter(frame == t + delta) %>% select(cell, state_next = state)
      if (nrow(now) == 0) next

      now <- now %>% left_join(nxt, by = "cell")
      out[[k]] <- data.frame(
        became_removed = as.numeric(now$state_next == 3),
        drug = d
      )
      k <- k + 1
    }
  }
  bind_rows(out)
}

trans_blocks <- build_transmission_blocks(tracks, delta)
recov_df     <- build_recovery_data(tracks, delta)

cat("Number of (frame, condition) transmission blocks:", length(trans_blocks), "\n")
cat("Number of recovery observations:", nrow(recov_df), "\n")

## -------------------------------------------------------------------------
## 3. Negative log-likelihood
## -------------------------------------------------------------------------
nll <- function(log_beta1, log_phi, betaD, log_gamma, gammaD) {

  beta1 <- exp(log_beta1)
  phi   <- exp(log_phi)
  gamma <- exp(log_gamma)

  ## --- transmission log-likelihood ---
  ll_trans <- 0
  for (blk in trans_blocks) {
    kappa_sum <- rowSums(exp(-blk$D / phi))
    lambda    <- beta1 * exp(betaD * blk$drug) * kappa_sum
    p_inf     <- 1 - exp(-lambda * delta)
    p_inf     <- pmin(pmax(p_inf, 1e-10), 1 - 1e-10)
    ll_trans  <- ll_trans + sum(dbinom(blk$became_infected, 1, p_inf, log = TRUE))
  }

  ## --- recovery log-likelihood ---
  p_rec <- 1 - exp(-gamma * exp(gammaD * recov_df$drug) * delta)
  p_rec <- pmin(pmax(p_rec, 1e-10), 1 - 1e-10)
  ll_rec <- sum(dbinom(recov_df$became_removed, 1, p_rec, log = TRUE))

  -(ll_trans + ll_rec)
}

## -------------------------------------------------------------------------
## 4. Fit
## -------------------------------------------------------------------------
## Reasonable starting values: eyeball beta1 from crude early growth rate,
## phi from grid spacing (e.g. ~ a few cell-widths), gamma from mean
## infectious-period length. Adjust these if optimizer struggles.
start_vals <- list(
  log_beta1 = log(0.01),
  log_phi   = log(3),
  betaD     = 0,
  log_gamma = log(0.02),
  gammaD    = 0
)

fit <- mle2(nll, start = start_vals, method = "BFGS",
            control = list(maxit = 500))

summary(fit)
ci <- confint(fit)
print(ci)

## -------------------------------------------------------------------------
## 5. Interpret results
## -------------------------------------------------------------------------
est <- coef(fit)

cat("\n--- Rate of infection ---\n")
cat("beta1 (baseline transmission rate, per frame):", exp(est["log_beta1"]), "\n")

cat("\n--- Spatial range of infection ---\n")
cat("phi (kernel decay distance, grid units):", exp(est["log_phi"]), "\n")

cat("\n--- Drug effect on transmission ---\n")
cat("betaD:", est["betaD"],
    " (exp(betaD) =", exp(est["betaD"]), "x change in transmission rate)\n")
cat("95% CI for betaD:", ci["betaD", ], "\n")
cat("--> CI excludes 0:", !(ci["betaD", 1] < 0 & ci["betaD", 2] > 0), "\n")

cat("\n--- Drug effect on recovery ---\n")
cat("gammaD:", est["gammaD"],
    " (exp(gammaD) =", exp(est["gammaD"]), "x change in recovery rate)\n")
cat("95% CI for gammaD:", ci["gammaD", ], "\n")

## -------------------------------------------------------------------------
## 6. Visual check: fitted kernel shape
## -------------------------------------------------------------------------
library(ggplot2)

d_seq <- seq(0, 20, length.out = 200)
kernel_df <- data.frame(
  d = d_seq,
  kappa = exp(-d_seq / exp(est["log_phi"]))
)

ggplot(kernel_df, aes(d, kappa)) +
  geom_line() +
  labs(x = "Distance between cells (grid units)",
       y = "Relative transmission weight",
       title = "Fitted spatial transmission kernel") +
  theme_minimal()

## -------------------------------------------------------------------------
## 7. Posterior predictive check: simulate and compare epidemic curves
## -------------------------------------------------------------------------
simulate_epidemic <- function(tracks_template, params, delta = 1) {
  # tracks_template: one condition's (cell, frame, x, y) skeleton, with
  # initial states at frame 1, used to re-simulate forward under the fit
  frames <- sort(unique(tracks_template$frame))
  state  <- tracks_template %>% filter(frame == frames[1]) %>%
    select(cell, x, y, state)

  sim_states <- list(state)

  for (t in frames[-length(frames)]) {
    coords_now <- tracks_template %>% filter(frame == t) %>% select(cell, x, y)
    state <- state %>% select(-x, -y) %>% left_join(coords_now, by = "cell")

    susc <- state %>% filter(state == 1)
    inf  <- state %>% filter(state == 2)

    if (nrow(susc) > 0 && nrow(inf) > 0) {
      D <- sqrt(outer(susc$x, inf$x, "-")^2 + outer(susc$y, inf$y, "-")^2)
      kappa_sum <- rowSums(exp(-D / params["phi"]))
      lambda <- params["beta1"] * exp(params["betaD"] * unique(tracks_template$drug)) * kappa_sum
      p_inf  <- 1 - exp(-lambda * delta)
      newly_infected <- susc$cell[runif(nrow(susc)) < p_inf]
      state$state[state$cell %in% newly_infected] <- 2
    }

    infd <- state %>% filter(state == 2)
    if (nrow(infd) > 0) {
      p_rec <- 1 - exp(-params["gamma"] * exp(params["gammaD"] * unique(tracks_template$drug)) * delta)
      newly_removed <- infd$cell[runif(nrow(infd)) < p_rec]
      state$state[state$cell %in% newly_removed] <- 3
    }

    sim_states[[length(sim_states) + 1]] <- state
    state$frame <- t + delta
  }

  bind_rows(sim_states)
}

# Example: compare observed vs simulated I(t) curve for one condition
# params <- c(beta1 = exp(est["log_beta1"]), phi = exp(est["log_phi"]),
#             betaD = est["betaD"], gamma = exp(est["log_gamma"]), gammaD = est["gammaD"])
# sim <- replicate(20, simulate_epidemic(df_before, params, delta), simplify = FALSE)
# (then plot mean +/- range of simulated I(t) against observed I(t))
```

```{r}
library(mcmc)

## -------------------------------------------------------------------------
## 1. Log-posterior = log-likelihood + log-prior
##    (weakly informative priors on the log/natural scale)
## -------------------------------------------------------------------------
log_posterior <- function(par) {
  log_beta1 <- par[1]; log_phi <- par[2]; betaD <- par[3]
  log_gamma <- par[4]; gammaD <- par[5]

  ll <- -nll(log_beta1, log_phi, betaD, log_gamma, gammaD)  # reuse existing nll

  # weakly informative priors: keeps parameters in a plausible range
  # without dominating the likelihood
  lp <- dnorm(log_beta1, mean = log(0.01), sd = 3, log = TRUE) +
        dnorm(log_phi,   mean = log(3),    sd = 3, log = TRUE) +
        dnorm(betaD,     mean = 0,         sd = 5, log = TRUE) +
        dnorm(log_gamma, mean = log(0.02), sd = 3, log = TRUE) +
        dnorm(gammaD,    mean = 0,         sd = 5, log = TRUE)

  ll + lp
}

## -------------------------------------------------------------------------
## 2. Initialize at the MLE, use its covariance to scale proposals
##    (this is the "easy" trick that makes plain Metropolis work well
##    without hand-tuning a proposal covariance from scratch)
## -------------------------------------------------------------------------
init <- coef(fit)                 # from the mle2 fit you already ran
Sigma <- vcov(fit)                # MLE covariance -> good proposal shape

## -------------------------------------------------------------------------
## 3. Run Metropolis-Hastings
## -------------------------------------------------------------------------
out <- metrop(log_posterior, initial = init, nbatch = 5000,
              scale = Sigma, blen = 1)

cat("Acceptance rate:", out$accept, "\n")   # aim for ~0.2-0.4; rescale if not

out2 <- metrop(out, scale = Sigma * 8)   # try a multiplier, adjust as needed
cat("Acceptance rate:", out2$accept, "\n")

# if still too high, scale up more; if it overshoots below ~0.2, scale back down
out3 <- metrop(out2, scale = Sigma * 8)
cat("Acceptance rate:", out3$accept, "\n")

out = out2
# if acceptance is too low/high, rerun with a scaled proposal, e.g.:
# out <- metrop(out, scale = Sigma * 0.5)   # shrink
# out <- metrop(out, scale = Sigma * 2)     # grow

## -------------------------------------------------------------------------
## 4. Posterior summaries
## -------------------------------------------------------------------------
post <- as.data.frame(out$batch)
colnames(post) <- c("log_beta1", "log_phi", "betaD", "log_gamma", "gammaD")

post$beta1 <- exp(post$log_beta1)
post$phi   <- exp(post$log_phi)
post$gamma <- exp(post$log_gamma)

burn <- 1:1000   # discard burn-in
post <- post[-burn, ]

summary(post[, c("beta1", "phi", "betaD", "gamma", "gammaD")])

# 95% credible intervals
apply(post[, c("beta1","phi","betaD","gamma","gammaD")], 2,
      quantile, probs = c(0.025, 0.975))
```




## Model Checking

```{r}
library(dplyr)

check_calibration <- function(trans_blocks, params, n_bins = 10) {
  all_p <- c(); all_y <- c()

  for (blk in trans_blocks) {
    kappa_sum <- rowSums(exp(-blk$D / params["phi"]))
    lambda <- params["beta1"] * exp(params["betaD"] * blk$drug) * kappa_sum
    p <- 1 - exp(-lambda * 1)  # delta = 1 frame
    all_p <- c(all_p, p)
    all_y <- c(all_y, blk$became_infected)
  }

  df <- data.frame(p = all_p, y = all_y) %>%
    mutate(bin = ntile(p, n_bins)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(p), mean_obs = mean(y), n = n())

  plot(df$mean_pred, df$mean_obs, xlim = c(0,1), ylim = c(0,1),
       xlab = "Mean predicted probability", ylab = "Observed frequency")
  abline(0, 1, col = "red")
  df
}

## Build the params vector from your mle2 fit
est <- coef(fit)
params <- c(
  beta1 = exp(est["log_beta1"]),
  phi   = exp(est["log_phi"]),
  betaD = est["betaD"]
)

## Run the calibration check on the transmission blocks you already built
calib_df <- check_calibration(trans_blocks, params, n_bins = 10)

## Inspect the table: mean_pred vs mean_obs should track closely per bin
print(calib_df)


check_calibration_recovery <- function(recov_df, params, n_bins = 10) {
  p <- 1 - exp(-params["gamma"] * exp(params["gammaD"] * recov_df$drug) * 1)

  df <- data.frame(p = p, y = recov_df$became_removed) %>%
    mutate(bin = ntile(p, n_bins)) %>%
    group_by(bin) %>%
    summarise(mean_pred = mean(p), mean_obs = mean(y), n = n())

  plot(df$mean_pred, df$mean_obs, xlim = c(0,1), ylim = c(0,1),
       xlab = "Mean predicted probability", ylab = "Observed frequency",
       main = "Recovery calibration")
  abline(0, 1, col = "red")
  df
}

params_recov <- c(gamma = exp(est["log_gamma"]), gammaD = est["gammaD"])
calib_recov_df <- check_calibration_recovery(recov_df, params_recov)

library(coda)

mcmc_obj <- as.mcmc(out$batch)
plot(mcmc_obj)             # trace + density plots — check for good mixing, no drift
effectiveSize(mcmc_obj)    # want this reasonably close to nbatch, not tiny
geweke.diag(mcmc_obj)      # |z| < ~2 for each parameter suggests convergence
```



```{r}
# Simulate many full epidemics under posterior draws (or MLE point estimate),
# using the simulate_epidemic() function from before
sim_curves <- function(tracks_template, post, n_sims = 50, delta = 1) {
  results <- list()
  for (s in 1:n_sims) {
    draw <- post[sample(nrow(post), 1), ]
    params <- c(beta1 = draw$beta1, phi = draw$phi,
                betaD = draw$betaD, gamma = draw$gamma, gammaD = draw$gammaD)
    sim <- simulate_epidemic(tracks_template, params, delta)
    curve <- sim %>% group_by(frame) %>%
      summarise(I = sum(state == 2), .groups = "drop")
    curve$sim_id <- s
    results[[s]] <- curve
  }
  bind_rows(results)
}

sims <- sim_curves(df_before, post, n_sims = 50)

obs <- df_before %>% group_by(frame) %>% summarise(I = sum(state == 2))

library(ggplot2)
ggplot() +
  geom_line(data = sims, aes(frame, I, group = sim_id), alpha = 0.1, color = "steelblue") +
  geom_line(data = obs, aes(frame, I), color = "black", linewidth = 1) +
  labs(title = "Observed (black) vs simulated (blue) I(t) curves")
```

```{r}
## -------------------------------------------------------------------------
## Simulate from a single fixed parameter set (the MLE), repeated n_sims times
## -------------------------------------------------------------------------
sim_curves_mle <- function(tracks_template, params, n_sims = 50, delta = 1) {
  results <- list()
  for (s in 1:n_sims) {
    sim <- simulate_epidemic(tracks_template, params, delta)
    curve <- sim %>% group_by(frame) %>%
      summarise(I = sum(state == 2), .groups = "drop")
    curve$sim_id <- s
    results[[s]] <- curve
  }
  bind_rows(results)
}

## Build the fixed parameter vector from your mle2 fit
est <- coef(fit)
params_mle <- c(
  beta1 = exp(est["log_beta1"]),
  phi   = exp(est["log_phi"]),
  betaD = est["betaD"],
  gamma = exp(est["log_gamma"]),
  gammaD = est["gammaD"]
)

## Run
sims <- sim_curves_mle(df_before, params_mle, n_sims = 50)

## Plot against observed, same as before
obs <- df_before %>% group_by(frame) %>% summarise(I = sum(state == 2))

library(ggplot2)
ggplot() +
  geom_line(data = sims, aes(frame, I, group = sim_id), alpha = 0.1, color = "steelblue") +
  geom_line(data = obs, aes(frame, I), color = "black", linewidth = 1) +
  labs(title = "Observed (black) vs simulated (blue) I(t) curves — MLE point estimate")
```


```{r}
sims <- sim_curves(df_after, post, n_sims = 50)

obs <- df_after %>% group_by(frame) %>% summarise(I = sum(state == 2))

library(ggplot2)
ggplot() +
  geom_line(data = sims, aes(frame, I, group = sim_id), alpha = 0.1, color = "steelblue") +
  geom_line(data = obs, aes(frame, I), color = "black", linewidth = 1) +
  labs(title = "Observed (black) vs simulated (blue) I(t) curves")


library(dplyr)

infectious_periods <- tracks %>%
  filter(state %in% c(2,3)) %>%
  group_by(cell, drug) %>%
  summarise(
    start = min(t_min[state == 2]),
    end   = min(t_min[state == 3], default = NA),
    .groups = "drop"
  ) %>%
  mutate(duration = end - start) %>%
  filter(!is.na(duration))

hist(infectious_periods$duration, breaks = 30)
mean(infectious_periods$duration); sd(infectious_periods$duration)
```


